---
title: "Sample Characterization"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2: 
    toc: FALSE
---

```{r sampleCharSetup, warning=FALSE, message=FALSE, echo = FALSE}
library(ProjectTemplate)
load.project() 

library(ggpubr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r sampleChar}
dada_ps <- readRDS("data/phyloseq_objects/dada_ps.rds")

pruned_ps <- sample_data(dada_ps) %>% data.frame() %>% 
    rownames_to_column(var = "sample_id") %>% 
    filter(t_fctr %in% c(0, 20)) %>% 
    .$sample_id %>% 
    prune_samples(dada_ps)

sample_data(pruned_ps)$t_fctr <- if_else(sample_data(pruned_ps)$t_fctr == 0, "Post","Pre")
sample_data(pruned_ps)$seq_run_id <- with(sample_data(pruned_ps), paste0(seq_lab, seq_run))
    
ord_plot_fun <- function(ord, ps){
    phyloseq::plot_ordination(ps, ordination = ord, 
                          shape = "seq_run_id", color = "t_fctr") + 
    facet_wrap(~biosample_id, ncol = 1) + theme_bw()
}

ord_bray <- ordinate(pruned_ps, method =  "PCoA", distance = "bray")
ord_jaccard <- ordinate(pruned_ps, method =  "PCoA", distance = "jaccard")
ord_unifrac <- ordinate(pruned_ps, method =  "PCoA", distance = "unifrac")
ord_wunifrac <- ordinate(pruned_ps, method =  "PCoA", distance = "wunifrac")
ord_plots <- list(bray = ord_bray, jaccard = ord_jaccard, 
                  unifrac = ord_unifrac, wunifrac = ord_wunifrac) %>% 
    map(ord_plot_fun, pruned_ps)
```

```{r sampleCharPlot, fig.cap = "A - Bray, B - Jaccard, C - Unifrac, D - Weighted unifrac. All plots showing unnormalized dada2 pipeline results."}
ggarrange(plotlist = ord_plots, ncol = 4, nrow = 1, 
          labels = "AUTO",
          legend = "bottom", common.legend = TRUE)
```

- Unmixed pre- and post-exposure samples are well separated for all individuals (Fig. \@ref(fig:sampleCharPlot)).     
- JHU 1 samples are distinct from other sequencing runs for unifrac across all individuals, though consistent difference between unmixed pre- and post-exposure samples.  
- Outlier PCR replicate for NIST runs  
- Pre- and post-expoure sample separation varies by individual.  
