---
title: "Bio v Tech Mixed Effects Analysis"
author: "Nate Olson"
date: "3/16/2018"
output: html_document
---

Using betadisper to calculate distance from centroid for each biosample then fitting a mixed effects model to characterize technical and biological variation components. 

__TODO__
* Write up summary for Hector and Nate to make sure approach is sound 
* Run on all data  
* Generate summary plot showing how variance components change by pipeline, normalization methods, and distance metrics. 

```{r setup, include=FALSE}
library(ProjectTemplate)
load.project()
library(vegan)
library(lme4)
library(broom) 
```

Functions for analysis 
```{r}
varcomp_lmer <- function(object){
    ## Modified code from HLMdiag::varcomp.mer to use group names
    vc <- lme4::VarCorr(fit)
    sig <- attr(vc, "sc")
    
    
    vc.mat <- bdiag(vc)
    vc.vec <- diag(vc.mat)
    vc.names <- names(vc) 
    
    res <- c(sig^2, vc.vec)
    names(res) <- c("sigma2", vc.names)
    return(res)
}


subset_dist <- function(dist_dat, sample_names){
    
    ## convert to matrix for subsetting
    dist_as_mat <- as.matrix(dist_dat) 
    
    ## subset distance matrix
    col_set <- colnames(dist_as_mat) %in% sample_names 
    row_set <- rownames(dist_as_mat) %in% sample_names 
    
    subset_mat <- dist_as_mat[row_set, col_set]
    
    ## return dist object 
    as.dist(subset_mat)
}
```

## Prepare data for analysis 
Example dataset DADA2 pipeline with Bray Curtis

```{r}
## Distance data
dada_bray <- readRDS("data/diversity_data/dada_RAW_bray.rds")
dist_obj <- dada_bray$result 


## Sample metadata
mgtstBiosample <- mgtstMetadata %>% 
    filter(biosample_id != "NTC", t_fctr %in% c(0,20)) 

mgtstBiosample <- mgtstBiosample %>% 
    mutate(col = str_remove(pos, "[:alpha:]"),
           col = as.numeric(col)) %>% 
    mutate(plate_half = if_else(col < 7, "A","B"),
           pcr_plate_half = paste(pcr_16S_plate, plate_half, sep = "_"),
           pcr_plate_half = factor(pcr_plate_half))

dist_obj <- subset_dist(dist_obj, mgtstBiosample$sample_id)

## Calculating sample distance from dataset centroid
betadispresults <- betadisper(dist_obj, 
                              rep(1, nrow(mgtstBiosample)), 
                              type = "centroid") 
mgtstBiosample$betadisp_dist <- betadispresults$distances

## Fitting mixed effects linear model
fit <- lmer(betadisp_dist ~ 1 + (1 | biosample_id / t_fctr ) + 
                (1  | seq_lab/ seq_run / pcr_16S_plate/ pcr_plate_half), 
            data = mgtstBiosample)

## Annotating sample data with model fit data
aug_df <- broom::augment(fit, mgtstBiosample)
```

Model fit statistics
```{r}
broom::glance(fit)
```

## Distances distribution
```{r}
mgtstBiosample %>% ggplot() + 
    ggridges::geom_density_ridges(aes(x = betadisp_dist, y = pcr_plate_half)) + 
    theme_bw() + 
    labs(x = "Distance from Centroid", y = "PCR Plate Half")
```


```{r}
ggplot(aug_df) + geom_point(aes(x = betadisp_dist, y = .resid)) + theme_bw() + 
    labs(x = "Distance from Centroid", y = "Residual")
```

```{r}
ggplot(aug_df) + 
    geom_qq(aes(sample = scale(.resid))) +
    geom_abline(aes(intercept = 0, slope = 1)) + 
    theme_bw() + 
    ggtitle("QQ-Plot")
```


```{r}
ggplot(aug_df) + 
    geom_point(aes(x = .fitted, y = .resid)) + 
    theme_bw() + 
    labs(x = "Fitted Values", y = "Residuals")
```

```{r}
data.frame(rand_var = varcomp_lmer(fit)) %>% 
    rownames_to_column(var = "group") %>% 
    mutate(varcent = rand_var/sum(rand_var) * 100,
           varcent = round(varcent, 2)) 
```


## Next steps
- apply method to all dist objects  
    - for each dist object 
        - subset dist  
        - get distances  
        - plot dist distribution 
        - fit model 
        - model diagnostic plots 
        - model stats 
        - generate varcomp_df 
- generate summary figures 
- identify conclusions 
- draft method description 