---
title: "Replicate Cluster Methods Development"
author: "Nate Olson"
date: "2/8/2018"
output: html_document
---
```{r message = FALSE, warning = FALSE, echo = FALSE}
library(cluster)
library(ProjectTemplate)
load.project()


cluster_eval_df <- list(beta_cluster_eval_bray_df, 
     beta_cluster_eval_jaccard_df, 
     beta_cluster_eval_unifrac_df, 
     beta_cluster_eval_wunifrac_df) %>% 
    map_df(tidy_cluster_eval)
```

## Objective  
The comparison to cluster results are noisier than expected, this could either be due to the noisy nature of real data, or clustering method finding local-maximum rather than optimal solution. 

```{r clusterEval, fig.cap = "Example comparison to expectation using clustering results."}
cluster_eval_df %>% 
    filter(seq_lab == "jhu", 
           seq_run == 2, 
           biosample_id == "E01JH0017",
           n_samples == 8,
           pipe %in% c("dada")) %>% 
    mutate(f_type = if_else(dist_method %in% c("unifrac_dist","wunifrac_dist"), 
                              "phylo","taxa"),
           weight = if_else(dist_method %in% c("unifrac_dist","jaccard_dist"),
                            "unweighted", "weighted")) %>% 
    ggplot() + 
    geom_line(aes(x = 2^-t_comp, 
                  y = cluster_results, 
                  color = method)) + 
    facet_grid(f_type~weight) + 
    theme_bw() + 
    labs(x = "Proportion Post-Exposure", 
         y = "Fraction Correct", color = "Normalization Method") + 
    theme(legend.position = "bottom")
```


## Approach
Using an example dataset, Bray-Curtis, for pipeline DADA2, individual E01JH0017, sequencing run JHU 2, rareified to 10000 sequences.   

1. Perform replicate partitioning around medioids with all possible combinations of seed medoids.
1. Look at variability of clustering results, can dive in more if necessary. 


## Method Development
```{r}
## Beta-diversity distance matrix
dist_obj <- readRDS("data/diversity_data/dada_rare10000_bray.rds")
dist_mat <- as.matrix(dist_obj$result)
```

```{r}
condensed_meta <- mgtstMetadata %>%
    filter(biosample_id != "NTC") %>%
    mutate(sample_id = as.character(sample_id)) %>%
    group_by(seq_lab, seq_run, biosample_id) %>%
    select(-pcr_16S_plate, -pos) %>%
    nest()

comparison_meta <- condensed_meta %>%
    select(-data) %>%
    add_column(t_comp = list(c(0:5, 10, 15))) %>%
    unnest() %>%
    left_join(condensed_meta)

full_comp_df <- comparison_meta %>%
    ## Generating data frame with sample_id and t_fctr for comparisons to perform
    mutate(comp_df = map2(data, t_comp, ~ filter(.x, t_fctr %in% c(.y, 20)))) %>%
    select(-data)
```

## Clustering Samples  
Mediouds and cluster assignments are all the same for the 28 potential medoid combinations. 

```{r}
comp_df <- full_comp_df$comp_df[[1]]
dist_mat <- as.matrix(dist_obj$result)
n_comp <- row.names(dist_mat) %in% comp_df$sample_id
sum(n_comp)
comp_mat <- dist_mat[row.names(dist_mat) %in% comp_df$sample_id,
                         colnames(dist_mat) %in% comp_df$sample_id]
cluster_combn <- combn(8,2, simplify = FALSE) %>% 
    map(~pam(x = comp_mat, medoids =  ., k = 2, diss = TRUE))
cluster_combn %>% map(pluck, "clustering") %>% 
    map_dfr(~data.frame(sample_id = names(.), 
                        cluster_assignment = .), 
            .id = "comb") %>% 
    spread(sample_id, cluster_assignment)
```

Applying the same method to all cluster assignments for full data set with bray rareified to 1000 reads. 


```{r}
eval_cluster_stability <- function(comp_df, dist_mat) {
    n_comp <- row.names(dist_mat) %in% comp_df$sample_id
    if (sum(n_comp) < 8) return(0)
    
    comp_mat <- dist_mat[row.names(dist_mat) %in% comp_df$sample_id,
    colnames(dist_mat) %in% comp_df$sample_id]
    
    cluster_combn <- combn(8, 2, simplify = FALSE) %>%
        map( ~ pam( x = comp_mat,medoids =  .,k = 2,diss = TRUE))
    
    cluster_combn %>% map(pluck, "clustering") %>%
        map_dfr( ~ data.frame(sample_id = names(.),
                              cluster_assignment = .),
                .id = "comb") %>%
    spread(sample_id, cluster_assignment) %>% 
    select(-comb) %>% distinct() %>% nrow()
}
```

```{r}
full_comp_df %>% 
    mutate(n_assignments = map_dbl(comp_df, eval_cluster_stability, dist_mat)) %>% 
    filter(n_assignments != 1)
```

```{r}
beta_df <- make_beta_div_df("bray")

get_stab_results <- function(dist_obj, full_comp_df){
    dist_mat <- as.matrix(dist_obj)
    full_comp_df %>% 
        mutate(n_assignments = map_dbl(comp_df, 
                                       eval_cluster_stability, 
                                       dist_mat))
}

assignment_eval_df <- beta_df %>% 
    mutate(dist_obj = map(dist_results, pluck, "result")) %>%
    filter(!is.null(dist_obj)) %>%
    mutate(assignment_output = map(dist_obj, get_stab_results, full_comp_df))
```

Most clustering results returned the same set of cluster assignments regardless of the medoids. 
```{r}
assignment_eval_df %>% 
    select(-dist_results, -dist_obj) %>% 
    unnest() %>% 
    ggplot() + geom_bar(aes(x = n_assignments))
```

```{r}
assignment_eval_df %>% 
    select(-dist_results, -dist_obj) %>% 
    unnest() %>% nrow()
```
600 out of 9600 assignments had multiple cluster results. 
A small number but enough to warrant inclusion including replicates. 
Need to figure out how to incorporate permutations into beta_eval_comparison pipeline. 

```{r}
assignment_eval_df %>% 
    select(-dist_results, -dist_obj) %>% 
    unnest() %>% 
    filter(n_assignments > 1) %>% 
    ggplot() + geom_bar(aes(x = n_assignments))
```


```{r}
assignment_eval_df %>% 
    select(-dist_results, -dist_obj) %>% 
    unnest() %>% 
    filter(n_assignments > 1) %>% 
    ggplot() + geom_bar(aes(x = biosample_id)) + 
    facet_grid(pipe~seq_lab*seq_run, scales = "free") + 
    theme_bw() + theme(axis.text.x = element_text(angle = 90))
```

```{r}
assignment_eval_df %>% 
    select(-dist_results, -dist_obj) %>% 
    unnest() %>% 
    filter(n_assignments > 1) %>% 
    group_by(pipe, biosample_id, seq_lab, seq_run, dist_method, method) %>% 
    summarise(count = n()) %>% 
    arrange(-count)
```

