---
title: "Ordination Plots"
author: "Nate Olson"
date: "5/7/2018"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggpubr)
library(phyloseq)
```

## Objective
Develop figure with ordination plots to evaluate whether TMM and RLE negatively impact beta diversity analysis. 

## Approach 
Generate ordination plots for data and use color to differentiate subjects and sequencing run (different plots). 
Compare normalization methods: RAW, RLE, TSS, and TMM for DADA2 and Mothur pipelines. 

```{r}
dada_tss_full <- readRDS("data/norm_data/dada_TSS.rds")
dada_tmm_full <- readRDS("data/norm_data/dada_TMM.rds")
mothur_tss_full <- readRDS("data/norm_data/mothur_TSS.rds")
mothur_tmm_full <- readRDS("data/norm_data/mothur_TMM.rds")
```

Subset samples only unmixed
create new variable names for subject titration 
seq run

```{r}
subset_ps <- function(ps){
    ps <- subset_samples(ps, t_fctr %in% c(20)) 
    filter_taxa(ps, function(x) sum(x) > 0, TRUE)
}

factor_var <- function(ps){
    sample_data(ps)$t_fctr <- factor(sample_data(ps)$t_fctr)
    sample_data(ps)$seq_run <- paste0(sample_data(ps)$seq_lab, 
                                      sample_data(ps)$seq_run)
    
    ps
}

prep_ps <- compose(subset_ps, factor_var)
```


```{r}
dada_ps_df <- tibble(pipe = "DADA2", 
                      normalization = c("TSS", "TMM")) %>% 
    add_column(ps = list(dada_tss_full, dada_tmm_full))

mothur_ps_df <- tibble(pipe = "Mothur", 
                      normalization = c("TSS", "TMM")) %>% 
    add_column(ps = list(mothur_tss_full, mothur_tmm_full))

make_ord_plot <- function(ps, ord, run_colors){
    plot_ordination(ps, ord, color = "seq_run", shape = "biosample_id") + 
        theme_bw()
}

ord_df <- bind_rows(dada_ps_df, mothur_ps_df) %>% 
    mutate(ps = map(ps, prep_ps)) %>% 
    mutate(Bray = map(ps, ordinate, method = "MDS"),
           `Weighted UniFrac` = map(ps, ordinate, method = "MDS", 
                                    distance = "wunifrac")) %>% 
    gather("dist_method", "ord", Bray, `Weighted UniFrac`)
    
ord_plot_df <- ord_df %>%     
    mutate(ord_plot = map2(ps, ord, make_ord_plot)) %>% 
    arrange(pipe, normalization, dist_method)
```

```{r fig.height = 8, fig.cap = "Principal coordinate analysis "}
ord_plot_dat_df <- ord_plot_df %>% 
    select(-ps, -ord) %>% 
    mutate(plot_dat = map(ord_plot, ~.$data),
           y_label = map_chr(ord_plot, ~.$labels$y),
           x_label = map_chr(ord_plot, ~.$labels$x)) %>% 
    mutate(x_label = str_extract(x_label, "(?<=\\[).*(?=\\])")) %>% 
    mutate(y_label = str_extract(y_label, "(?<=\\[).*(?=\\])")) %>% 
    mutate(dca_percent = paste0("DCA1: ", x_label, "\n","DCA2: ", y_label)) %>% 
    select(-ord_plot) %>% 
    unnest()

dada_plot <- ord_plot_dat_df %>% 
    filter(pipe == "DADA2") %>%
    ggplot() + geom_point(aes(x = DCA1, y = DCA2, color = seq_run, 
                              shape = biosample_id)) + 
    geom_text(aes(x = 1.5, y = -0.8, label = dca_percent)) + 
    facet_grid(normalization ~ dist_method) + 
    theme_bw() + 
    labs(color = "Sequencing Run", shape = "Subject") + 
    scale_color_manual(values = run_colors)

mothur_plot <- ord_plot_dat_df %>% 
    filter(pipe == "Mothur") %>% 
    ggplot() + 
    geom_point(aes(x = DCA1, y = DCA2, color = seq_run, shape = biosample_id)) +
    geom_text(aes(x = 0.8, y = -0.5, label = paste0("DCA1: ", x_label))) + 
    geom_text(aes(x = 0.8, y = -0.7, label = paste0("DCA2: ", y_label))) +
    facet_grid(normalization ~ dist_method) + 
    theme_bw() + 
    labs(color = "Sequencing Run", shape = "Subject") + 
    scale_color_manual(values = run_colors)

ggarrange(dada_plot + rremove("xlab"), mothur_plot, 
          ncol = 1,nrow = 2, labels = "AUTO",
          common.legend = TRUE, legend = "bottom") 
```

