---
title: "Bio v. Tech Analysis"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::html_document2: 
    toc: FALSE
---
```{r}
library(ProjectTemplate)
load.project()
library(ggfortify)
library(ggthemes)
library(stringr)
source("src/plot_color_definitions.R")
```

## Comparison of biological and technical factors 

Figure 

```{r}
bio_v_tech_dat <- biol_v_tech_variation[[3]] %>% 
    filter(normalization == "RAW") #%>% 
    #filter(!grepl("jhu_run1", sample_a), !grepl("jhu_run1", sample_b))
```


Excluding JHU Run 1, low quality and high variability. 
```{r fig.height = 5}
bio_v_tech_dat %>% 
    ggplot(aes(x = variation_label, y = value, color = variation, fill = variation)) + 
    geom_tufteboxplot() + 
    facet_grid(pipe~metric) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = -45, hjust = 0), 
          legend.position = "bottom") + 
    labs(x = "Variation Type", y = "Beta Diversity", 
         fill = "Variation Source", color = "Variation Source")
```


### Fitting linear model

```{r}
bio_v_tech_fit <- bio_v_tech_dat %>% 
    rename(dist_method = metric) %>% 
    group_by(dist_method, pipe) %>% 
    nest() %>% 
    mutate(fit = map(data, ~lm(log(value) ~ variation + variation_label, data = .))) 

bio_v_tech_anova <- bio_v_tech_fit %>% 
    mutate(anova_fit = map(fit, anova), 
           tidy_anova = map(anova_fit, broom::tidy)) %>% 
    unnest(tidy_anova) 

bio_v_tech_tukey <- bio_v_tech_fit %>% 
    mutate(anova_fit = map(fit, aov), 
           tukey_fit = map(anova_fit, TukeyHSD),
           tidy_tukey = map(tukey_fit, broom::tidy)) %>% 
    unnest(tidy_tukey) 
```


### Model Results
Seq runs are significantly different for all pipelines/ metrics 
```{r}
bio_v_tech_anova  %>% knitr::kable()
```

```{r fig.height = 6}
comp1_levels <- c("biological","technical",
                  "btw_labs","w/in_lab_runs", 
                  "w/in_lab_pcr","btw_time", 
                  "w/in_subj_btw_time")

comp2_levels <- c("biological","technical",
                "btw_subj_w/in_time", "btw_labs", "w/in_subj_btw_time", 
                "btw_time", "w/in_lab_pcr", "w/in_lab_runs")

bio_v_tech_tukey %>% 
    mutate(adj.p.value = adj.p.value * 384) %>% 
    mutate(adj.p.value = if_else(adj.p.value > 0.06, 0.06, adj.p.value)) %>%
    mutate(pos_neg = if_else(estimate > 0, "comp1>comp2", "comp2>comp1")) %>%
    separate(comparison, c("comp1","comp2"), sep = "-",remove = FALSE) %>% 
    mutate(comp1 = factor(comp1, levels = comp1_levels)) %>%
    mutate(comp2 = factor(comp2, levels = comp2_levels)) %>%
    mutate(estimate = if_else(adj.p.value < 0.05, estimate, NaN)) %>%
    filter(comparison != "technical-biological") %>%
    ggplot() + 
    geom_raster(aes(x = comp1, y = comp2, fill = adj.p.value)) + 
    geom_text(aes(x = comp1, y = comp2, label = round(estimate,2), color = pos_neg)) + 
    facet_grid(pipe~dist_method) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = -45, hjust = 0),
          legend.position = "bottom") + 
    scale_fill_gradient(low = "black", high = "grey90")
```

```{r}
bio_v_tech_tukey
```


```{r}
bio_v_tech_tukey %>% 
    mutate(adj.p.value = adj.p.value * 384) %>% 
    mutate(pos_neg = if_else(estimate > 0, "comp1>comp2", "comp1<comp2")) %>% 
    # separate(comparison, c("comp1","comp2"), sep = "-",remove = FALSE) %>% 
    # mutate(comparison = str_remove_all(comparison, "ist_run|hu_run")) %>% 
    # mutate(estimate = if_else(adj.p.value < 0.05, estimate, NaN)) %>% 
    filter(comparison == "technical-biological") %>%
    ggplot() + 
    geom_raster(aes(x = comparison, y = pipe, fill = log10(adj.p.value))) + 
    geom_text(aes(x = comparison, y = pipe, label = round(estimate,2), color = pos_neg)) + 
    facet_grid(.~dist_method) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = -45, hjust = 0),
          legend.position = "bottom")
```

__NOTES__  
_Methods_  
* Fit independent linear models to d~value ~ variation + variation_label for each pipeline and metric.  
* Used TukeyHSD to test for pairwise distances between runs.  

_Results_  
*  For raw count data
    - biological factors are significantly greater source of variation than technical sources for all pipeline and diversity metrics. 
        + Magnitude of difference varies by pipeline and metric, though general trends oberved
    - For individual sources of variation
        + between labs largest individual source of variation 
            * excluding within subject bwetween time for wunif dada 
            * No significante for all 
        + Followed by within lab runs and between subject within time 
            - No significant difference for deblur (Unifrac and jaccard) Denovo (wunifrac)
    - PCR replicates smallest pairwise distance for nearly all pipelines and metrics
        + Significant for most 
* For deblur and Unweighted metrics nad qiime de novo weighted unifrac beta diversity between different factors were not significantly different though observed trends were consistent with other pipelines and metrics. 

_Interpretation_   
* Combined sources of biological variation had greater beta diversity between samples compared to technical sources of variation.
* Individual technical sources of variation were often greater than individual biological sources of varation. 
* Lack of significant difference - due to feature and data characteristics (high rate of phylogenetically closesly related features for qiime de novo), not sure about Deblur intrepretation. 


_Potential analysis improvements_   
* Better presentation of Tukey results
* Using lm bio v. tech results for bio - tech plot  
* Better model fit?

### Raw Stats
```{r}
bio_v_tech_fit %>% mutate(aov_fit = map(fit, aov), 
                   tidy_aov = map(aov_fit, broom::tidy)) %>% 
    unnest(tidy_aov) %>% 
    mutate(sumsq = round(sumsq, 3),
           meansq = round(meansq, 3),
           statistic = round(statistic, 1)) %>% 
    knitr::kable()
```

### Fit Summary Statistics 
```{r}
bio_v_tech_fit %>% mutate(glance_fit = map(fit, broom::glance)) %>% 
    unnest(glance_fit) %>% 
    select(-data, -fit) %>% 
    gather(metric, value, -dist_method, -pipe) %>% 
    spread(dist_method, value) %>% 
    filter(metric %in% c("adj.r.squared","AIC","p.value")) %>% 
    knitr::kable(digits = 3)
```

### Diagnostic Plots
```{r}
bio_v_tech_fit %>% 
    mutate(diag_plots = map2(fit, paste(pipe, dist_method), ~{autoplot(.x) + ggtitle(.y)})) %>% 
    {print(.$diag_plots)}
```


## Norm impact
Using varpart compare variation in data attributed to different biological and technical factors.
Evaluate stats - lack of or singificance when normalized relative to raw. 

```{r bioVtechPlot2, warning=FALSE, message=FALSE, echo = FALSE, fig.cap = "Biological vs. Technical Variation, y-axis is the adjusted R^2 value, indicating proportion of variance explained by each biological (subject and titration) and technical (seq run) variable. Black bars represent values for un-normalized data."}
# norm_method_set <- c("RAW", "rare5000", "rareq15", "RLE", "TMM")
varpart_trim <- varpart_stats %>% 
    filter(effect == "conditional", 
           # normalization %in% norm_method_set, 
           metric %in% c("bray", "wunifrac")) #%>%
    # mutate(normalization = factor(normalization, 
    #                               levels = norm_method_set, 
    #                               ordered = T))

varpart_trim_raw <- varpart_trim %>% filter(normalization == "RAW")
varpart_trim_norm <- varpart_trim %>% filter(normalization != "RAW")

ggplot(varpart_trim_norm) +
    geom_bar(aes(x = metric, y = Adj.R.square, fill = normalization),
             stat = "identity", position = "dodge", width = 0.5) +
    geom_point(data = varpart_trim_raw,
               aes(x = metric, y = Adj.R.square, color = significance), 
shape = "-", size = 8) +
    facet_grid(feature~pipe, scales = "free_y", space = "free") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
    xlab("") + ylab("Adjusted R^2") +
    theme(legend.position = "bottom") + 
    scale_fill_manual(values = norm_colors)

``` 

__NOTES__  
_Methods_  
* Used variation partitioning to estimate the proportion of total observed variation due to sequencing run, biological replicates, and time points. 
 
_Results_  
* All rarefaction levels reduced variation attributed to sequencing runs
    - No longer significant for ??? 
* Increase in seq run variation after RLE and TMM
* Rareq15, RLE, and TMM decrease bio for some
* CSS and TSS - minimal to no change 


_Interpretation_   
* 


_Potential analysis improvements_   
* Better figures 
* p value multiple comparisons correction

```{r}
# norm_method_set <- c("RAW", "rare5000", "rareq15", "RLE", "TMM")
varpart_trim <- varpart_stats %>% 
    filter(effect == "conditional", 
           # normalization %in% norm_method_set, 
           metric %in% c("jaccard", "unifrac"))# %>%
    # mutate(normalization = factor(normalization, 
    #                               levels = norm_method_set, 
    #                               ordered = T))

varpart_trim_raw <- varpart_trim %>% filter(normalization == "RAW")
varpart_trim_norm <- varpart_trim %>% filter(normalization != "RAW")

ggplot(varpart_trim_norm) +
    geom_bar(aes(x = metric, y = Adj.R.square, fill = normalization, color = significance),
             stat = "identity", position = "dodge", width = 0.5) +
    geom_point(data = varpart_trim_raw,
               aes(x = metric, y = Adj.R.square, color = significance), 
               shape = "-", size = 8) +
    facet_grid(pipe~feature, scales = "free_x", space = "free") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
    xlab("") + ylab("Adjusted R^2") +
    theme(legend.position = "bottom") + 
    scale_fill_manual(values = norm_colors)
```

__NOTES__  
_Methods_  
* Used variation partitioning to estimate the proportion of total observed variation due to sequencing run, biological replicates, and time points. 
 
_Results_  
* All rarefaction levels reduced variation attributed to sequencing runs
    - No longer significant for deblur and mothur 
    - Only still significant source for variation for data
* All rarefaction levels saw increased variation attributed to subject.
* Minimal change titration 
similar results for the two unweighted metrics
* Closed reference, open reference, and de novo subject variation greater than seq run after normalization but not without. 


_Interpretation_   
* 


_Potential analysis improvements_   
* Better figures 
* p value multiple comparisons correction


## Norm Impact no jhu1
```{r bioVtechPlot2noJHU, warning=FALSE, message=FALSE, echo = FALSE, fig.cap = "Biological vs. Technical Variation, y-axis is the adjusted R^2 value, indicating proportion of variance explained by each biological (subject and titration) and technical (seq run) variable. Black bars represent values for un-normalized data."}
# norm_method_set <- c("RAW", "rare5000", "rareq15", "RLE", "TMM")
varpart_trim <- varpart_stats_no_jhu1 %>% 
    filter(effect == "conditional", 
           # normalization %in% norm_method_set, 
           metric %in% c("bray", "wunifrac")) #%>%
    # mutate(normalization = factor(normalization, 
    #                               levels = norm_method_set, 
    #                               ordered = T))

varpart_trim_raw <- varpart_trim %>% filter(normalization == "RAW")
varpart_trim_norm <- varpart_trim %>% filter(normalization != "RAW")

ggplot(varpart_trim_norm) +
    geom_bar(aes(x = metric, y = Adj.R.square, fill = normalization),
             stat = "identity", position = "dodge", width = 0.5) +
    geom_point(data = varpart_trim_raw,
               aes(x = metric, y = Adj.R.square, color = significance), 
shape = "-", size = 8) +
    facet_grid(feature~pipe, scales = "free_y", space = "free") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
    xlab("") + ylab("Adjusted R^2") +
    theme(legend.position = "bottom") + 
    scale_fill_manual(values = norm_colors)

``` 


```{r}
# norm_method_set <- c("RAW", "rare5000", "rareq15", "RLE", "TMM")
varpart_trim <- varpart_stats_no_jhu1 %>% 
    filter(effect == "conditional", 
           # normalization %in% norm_method_set, 
           metric %in% c("jaccard", "unifrac"))# %>%
    # mutate(normalization = factor(normalization, 
    #                               levels = norm_method_set, 
    #                               ordered = T))

varpart_trim_raw <- varpart_trim %>% filter(normalization == "RAW")
varpart_trim_norm <- varpart_trim %>% filter(normalization != "RAW")

ggplot(varpart_trim_norm) +
    geom_bar(aes(x = metric, y = Adj.R.square, fill = normalization, color = significance),
             stat = "identity", position = "dodge", width = 0.5) +
    geom_point(data = varpart_trim_raw,
               aes(x = metric, y = Adj.R.square, color = significance), 
               shape = "-", size = 8) +
    facet_grid(pipe~feature, scales = "free_x", space = "free") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
    xlab("") + ylab("Adjusted R^2") +
    theme(legend.position = "bottom") + 
    scale_fill_manual(values = norm_colors)
```