---
title: "Seq char and qual figures and stats"
author: "Nate Olson"
date: "2/15/2018"
output: html_document
---

```{r setup, include=FALSE}
library(ProjectTemplate)
load.project()
```

```{r}
qual_df <- seq_qual_comparisons %>% 
    mutate(read_dir = paste0(read_dir, "_qual")) %>% 
    select(-sd_quality, -cov_quality) %>% 
    spread(read_dir, mean_quality) %>% 
    separate(replicate, c("biosample_id","t_fctr","seq_run_id"), remove = FALSE,extra = "merge")

tech_art_df <- seq_char_comparisons %>% 
    select(replicate,  pipe, normalization, metric, cov_num_reads, cov_total_abu) %>% 
    left_join(qual_df)

tech_info_df <- tech_art_df %>% 
    filter(normalization == "RAW") %>% 
    select(biosample_id, t_fctr, seq_run_id, pipe, cov_total_abu, R1_qual, R2_qual) %>% 
    distinct() 

```

## Total Abundance COV

COV by pipeline and sequencing run
```{r}
tech_art_df %>% 
    mutate(seq_run_id = factor(seq_run_id)) %>% 
    ggplot() + 
    geom_boxplot(aes(x = pipe, y = cov_total_abu, color = seq_run_id)) + 
    theme_bw()
```

COV by pipeline and seqencing run showing relationship between PCR sets across sequencing runs
```{r}
tech_art_df %>% 
    mutate(seq_run_id = factor(seq_run_id)) %>% 
    ggplot() + 
    geom_point(aes(x = seq_run_id, y = cov_total_abu)) + 
    geom_line(aes(x = as.numeric(seq_run_id), y = cov_total_abu, group = paste(biosample_id, t_fctr))) + facet_wrap(~pipe) + theme_bw()
```




Testing for differences using a mixed effects model
```{r}
fit2 <- lme4::lmer(cov_total_abu ~ seq_run_id + (1 | pipe / biosample_id / t_fctr), 
                   data = tech_art_df %>% mutate(pcr_set = paste(pipe, biosample_id, t_fctr)))
```

```{r}
summary(fit2)
```

Not ideal fit but results show differences between the first and second runs at the two sequencing centers. 
```{r}
plot(fit2,type=c("p","smooth"))
```

```{r}
plot(fit2,sqrt(abs(resid(.)))~fitted(.), type = c("p","smooth"))
```


```{r}
lattice::qqmath(fit2)
```

```{r}
confint_dat <- confint(fit2)
```

```{r}
fit_coef <- coef(fit2)
fit_coef_tidy <- fit_coef$pipe %>% rownames_to_column(var = "pipe") %>% 
    gather("run_id","cov_est", -pipe, -`(Intercept)`) %>% 
    mutate(cov_est = `(Intercept)` + cov_est) %>% 
    mutate(run_id = str_replace(run_id, "seq_run_id","")) %>% 
    rename(jhu_run1 = `(Intercept)`) %>%
    spread(run_id, cov_est) %>%
    gather("run_id","cov_est", -pipe)
```

```{r}
fit_coef_tidy %>% ggplot() + geom_point(aes(x = pipe, y = cov_est, color = run_id)) + theme_bw()
```

```{r}
mod_betas <- fit2@beta
cov_est <- data_frame(run_id = c("jhu_run1","jhu_run2","nist_run1","nist_run2"),
                      est = c(mod_betas[1], 
                              mod_betas[1] + mod_betas[2],
                              mod_betas[1] + mod_betas[3],
                              mod_betas[1] + mod_betas[4]))
                    
fit_est_df <- as.data.frame(confint_dat) %>% 
    rownames_to_column(var = "run_id") %>% 
    filter(!grepl("sig", run_id)) %>% 
    gather("conf_int","est", -run_id) %>% 
    spread(run_id, est) %>%
    rename(jhu_run1 = `(Intercept)`) %>%
    gather("run_id","est", -conf_int, -jhu_run1) %>% 
    mutate(est = jhu_run1 + est) %>% 
    mutate(run_id = str_replace(run_id, "seq_run_id","")) %>%
    spread(run_id, est) %>% 
    gather("run_id","est", -conf_int) %>% 
    mutate(conf_int = if_else(conf_int == "2.5 %", "lci","uci")) %>% 
    spread(conf_int, est) %>% 
    left_join(cov_est)
```


```{r}
ggplot(fit_est_df) + 
    geom_pointrange(aes(x = run_id, y = est, ymin = lci, ymax = uci)) + 
    labs(x = "Sequencing Run", y = "Total Abundance COV") + theme_bw()
```

## Sequnce Quality Score Analysis  
```{r}
seq_qual_dat <- seq_char_df %>% 
    left_join(mgtstMetadata) %>% 
    filter(biosample_id != "NTC") %>% 
    mutate(seq_run_id = paste0(seq_lab, seq_run))

seq_qual_dat %>% 
    ggplot() + 
    geom_boxplot(aes(x = seq_run_id, y = quality, color = read_dir))  + theme_bw() + 
    labs(x = "Sequencing Run", y = "Mode Read Base Quality Score", color = "Read Direction") + 
    theme(legend.position = "bottom")
```

Testing for statistical difference in read quality between sequencing runs. 
```{r}
fit <- lm(quality~ - 1 + read_dir*seq_run_id, data = seq_qual_dat)
```

```{r}
plot(fit)
```

```{r}
summary(fit)
```
```{r}
anova(fit)
```
```{r}
aov(fit) %>% TukeyHSD()
```


